Data Management \\ and LSST Special Programs

### 2017-11-10 ms.tex
major updates to Action Items

\author{M.~L.~Graham, M.~Juri\'{c}, and E.~Bellm}

@@

\setDocAbstract{The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community. To do this, we begin with a review of the broad science goals of the previously proposed Special Programs and the potential diversity of data they might generate. We also consider any technical limitations on the data from the telescope, camera, and/or DM processing capabilities, and the methods by which information could be exchanged with the scientific community on these issues (i.e., the upcoming call for white paper proposals). We then move on to summarize the current DM plans for processing Special Programs data, and highlight specific requirements or text in the existing documentation that might need additions or clarifications to reflect these plans. We also develop a series of case studies for DM's processing of Special Programs data to better illuminate any complex or hidden issues; these case studies could also serve as examples for the upcoming call for community white papers.}

\setDocChangeRecord{%
\addtohist{1}{2017-04-??}{Internal working document.}{Melissa Graham}
%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
}

@@

The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):

@@

\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent. In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans.

\subsection{Study Overview}\label{ssec:intro_overview}

Section \ref{sec:data} compiles a comprehensive overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters. We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera, the scientific needs of as many currently proposed Special Programs as we can find, and the future mechanisms for LSST to gather more community input. We also consider whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.

Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data: reconfiguring pipelines to create unique products, incorporating data into the WFD main survey products, and enabling end-user Level 3 processing. Having established the {\it intent} of DM, we step through each of the relevant LSST documents and raise {\bf "Concerns"}, which highlight specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data. These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues. {\bf The Concerns raised by this study fall into two main categories:}
\begin{enumerate}
\item Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be.
\item Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD.
\end{enumerate}

Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys. These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional {\bf Concerns} for the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.

@@

To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data. The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (standard visit images), then it might not be adequately handled by the DM system. To get a handle on the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes (Section \ref{ssec:data_bounds}). Next, we explore the current set of proposed science programs for LSST (Section \ref{ssec:data_science}) and the types of observations they would want to do, and then describe when and how we can incorporate revisions to these proposals from the user community in the future (Section \ref{ssec:data_comm}). Finally, we explore whether DM can or should set down any technical constraints on how far an observing program can deviate from a sequence of standard visit images and still be processed by DM pipelines, be they the reconfigured and/or the standard Level 1 and 2 pipelines (Section \ref{ssec:data_DMconstraints}).

% % % % % % % % % % % % % % % % % % 
\subsection{Observing Boundaries for Special Programs}\label{ssec:data_bounds}

We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals. So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}. \cite{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data). Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \cite{2008arXiv0805.2366I}). We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.

>>>>

\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, and E.~Bellm}

@@

\setDocAbstract{{\bf This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget. Unresolved issues spawn {\bf Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}

\setDocChangeRecord{%
\addtohist{1}{2017-11-??}{Internal working document.}{Melissa Graham}
%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
}

@@

The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):

@@

\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated. This study spawns {\bf Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.

%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an {\bf Action} is spawned.

Section \ref{sec:data} compiles an overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters. We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera.

Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data, including the capabilities for processing non-standard visits and whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes. We also describe DM's intent for incorporating Special Programs data into the WFD main survey products, reconfiguring pipelines to create unique products, and providing support for end-user Level 3 processing.

Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
%These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues. {\bf The Concerns raised by this study fall into two main categories:} \\
%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
%2. Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget. \\

Section \ref{sec:SP_SC} gathers the existing efforts from the Science Community regarding Special Programs, summarizes the scientific needs of as many currently proposed Special Programs as we can find, and outlines the future mechanisms for LSST to gather more community input -- including some recommendations for the future call for Special Programs white paper proposals.

Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys. These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional issues regarding the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.

@@

To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data. The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (i.e., standard visit images), then it might not be adequately handled by the DM system. To get an idea of the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes, in Section \ref{ssec:data_bounds}.

We can also get an idea of the expected diversity in exposure type and survey pattern from the current set of proposed Special Programs for LSST, which are reviewed in more detail in Section \ref{sec:SP_SC}. We find that most of the proposed Special Programs are likely to use standard visits, two $\sim15$ second exposures, and that the potential diversity of Special Programs data is mainly comprised of exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, long series of exposures obtained of the same field (i.e., deep drilling), and images of crowded fields.

All aspects related to DM's {\it processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.

% % % % % % % % % % % % % % % % % % 
\subsection{Technical Boundaries for Special Programs Data}\label{ssec:data_bounds}

We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals. For the purposes of this document we consider the ``observing boundaries" as those imposed by the camera, telescope, or site, and independent of any DM processing capabilities. It is not the job of this document to define these observing boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.

The current technical limitations on what kind of observations can be obtained with the LSST telescope and camera are described below in a rough draft, to be updated in the future. Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.

Filter Changes: \\
$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293). \\
$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances. However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes. \\
$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights. \\

Exposure Times: \\
$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
$\bullet$ The maximum exposure time is not restricted. \\
$\bullet$ However, for exposure times there are other considerations. The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question. Changing the exposure time also affects the photometric and astrometric calibrations. Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag. A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations. Additionally, the impact on CR rejection routines is untested for long exposures.

Camera Rotation: \\
$\bullet$ The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields? \textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism? Can't remember where I heard it.}

Total Time: \\
$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey). This could be increased a bit, but otherwise imposes an upper limit on the total time request of a given Special Program.

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{Bounds-1} The purpose of this ticket is to request clarification regarding several aspects of the telescope and camera capabilities that might impose boundaries on the diversity of data from Special Programs. First, have the following proposed filter change tolerances been confirmed: (1) the minimum time between filter changes has no restrictions from thermal tolerances, but it is recommended to keep the filter change rate lower than once every 10 minutes in order to minimize overheads; (2) the maximum total number of filter changes is 100000 over 15 years (18/night) and the maximum number of filter swaps in and out of the carousel is 3000 in 15 years (once every two nights). Second, are there any restrictions on the use of the camera rotator or the total amount of slew time? For example, could a Special Program continuously jump between well-separated fields on the sky for an extended amount of time? \end{enumerate}

As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (DPDD, LSE-163, \cite{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).

In this section we review in greater detail the planned ability of DM to process non-standard visit images (Section \ref{ssec:dmplans_NSV}), the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).

% % % % % % % % % % % % % % % % % % 
\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:dmplans_NSV}

A standard visit image is considered to be two snap exposures of $\sim15$ second each, and in the WFD main survey sequential visits will be of different fields. As we found in Section \ref{sec:data}, Special Programs are most likely to request observing modes with shorter exposure times and long sequences of visits to the same field. In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).

\subsubsection{Exposure Times}

Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD. We consider shorter and longer exposure in turn.

{\bf Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds). The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

{\bf Longer Exposures.} There is no maximum exposure time specified for an LSST image. Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine. However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources. With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).

Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them? This is probably best left ot a policy issue (beyond the scope of this document).

\subsubsection{Image Acquisition Rate}

OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds. However, {\it if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated {\tt DIASources}.

{\bf Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds. This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later? {\bf How long does it take to save an image to disk?}

{\bf Processing backup:} Are there options for short-term increases in parallel processing power at NCSA? (Such options might be needed anyway to process crowded fields with $>$10k {\tt DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load. However, at NCSA there should be ways to elastically change the amount of processing power available. This is also an issue for crowded fields (below).

{\bf Unassociated {\tt DIASources}:} the consistency of {\tt DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field). In other words, it is not guaranteed that the next string of processing would have access to the latest {\tt DIAObjects} for association. It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations. This is also an issue for crowded fields (below).

\subsubsection{Exposures per Visit, and Long Sequences of a Single Field}

K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. {\it MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

K.-T. has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler. One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence. The processing pipeline would need to know how to divide this sequence up into visits. As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.

\subsubsection{Alerts from Crowded Fields}

Special Programs are more likely to include crowded fields than the WFD main survey area. Due to the increased number of sources, the number of {\tt DIASources} -- and therefore the number of Alerts -- increases as well. In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production. A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete. K.-T. reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed. This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).

\subsubsection{Summary}

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{NSV-1} The purpose of this ticket is to ensure that DM's capability for processing non-standard visit images, and required by LSR-REQ-0111, will meet the needs of Special Programs data (which might be considerably more diverse than data for the wide-fast-deep main survey). For now we have adopted a broad interpretation of ``processing" to mean both the initial ISR stage and also the Level 1 Alert Production pipeline, because allowing as many LSST images as possible to contribute to the Alert Stream is a benefit to time domain science. After comparing the anticipated potential diversity in Special Programs data with the currently planned DM pipelines, we have identified several open questions that need to be resolved in advance of the call for Special Programs white papers in the spring of 2018. (1) What is the anticipated minimum (and maximum, if applicable) exposure time for an image to be successfully processed by ISR and DIA? In the latter case, this will probably be limited by the time needed to generate a well-formed PSF. (2) Presently, OSS-REQ-0291 requires that even short exposures be spaced by 15 seconds. Might this requirement be slackened to accommodate a higher cadence Special Program? If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping {\it vs.} delaying Alerts during observations with a fast ingest cadence. These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts. Alternatively, we could formally require that only 30 second visits will be allowed to contribute to Level 1 Alert Production (3) Long sequences of images on a single field are a likely Special Programs mode (e.g., DDF). We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits. Consequential images of the same field may lead to unassociated {\tt DIASources} of the same {\tt DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately. \end{enumerate}

DM will want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science. In the following three sections we summarize when and how Special Programs data might be incorporated into the pipelines and products of the Difference Imaging Analysis and Alert Stream, the Data Release Pipeline products, and/or MOPS.

It is beneficial to transient science to include as many LSST images into the Level 1 Difference Imaging Analysis and Alert Stream as possible. Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST (this prospect is discussed further in Section \ref{ssec:dmplans_L3}). Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream. DM's ability to process the potential diversity of image types from Special Programs is explored above in Section \ref{ssec:dmplans_NSV}, and we continue the discussion of incorporating processed images into the Level 1 DIA pipeline.

{\bf Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template. So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine. When processing Special Programs data with the Level 1 pipeline, certain science cases might call for the capability to load and use a certain template that is different from the WFD template (i.e., built over a different timescale). K.-T. confirms that there is not enough memory allotted to store more than one template over the whole sky, but for sub-regions, storing and using an alternative template should be possible. This is not an issue limited to Special Programs, since during commissioning it is conceivable that multiple template versions. K.-T. also confirms that the capability for the processing pipeline to choose a given template based on the programID in the raw image metadata will exist.

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries). Since MOPS takes {\tt DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS. As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway. There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.

Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible. We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence. When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP. For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions. In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey. This is mostly discussed here in context with the {\tt Source} and {\tt Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.

LSST DM intends to reconfigure its existing pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible. In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.

In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program. For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}). This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.

The current outstanding issue related to DM's processing of special programs data is that the work-hours needed to reconfigure and test the pipelines, run them and verify the data products for public release -- potentially on intermediate timescales that do not coincide with the prompt/yearly Level 1 and 2 timescales, e.g., monthly stacks of deep drilling fields -- have not been included in the personnel budget and thus, at the moment, constitute an expansion of scope. As we describe in Section \ref{ssec:docrev_dmsr} there are requirements in the DMSR that state that DM does have the responsibility to process Special Programs data.

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-1} The purpose of this ticket is to investigate the DM staffing needs required to reconfigure pipelines for Special Programs, test and validate them, and then run and verify their products -- during commissioning and into operations. Questions to be answered include, e.g., product ownership of the reconfigured pipelines and test data products, and how do we set goals and deadlines for this? \end{enumerate}

As described in Section \ref{ssec:docrev_dmsr}, the DMSR contains requirement DMS-REQ-0320, which states that "it shall be possible for special programs to trigger their own data processing recipes". Assuming that this means that a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline, it would appear that auto-triggering of alternative reduction pipelines is not currently supported by the DM system.

Also described in Section \ref{ssec:docrev_dmsr} is the DMSR requirement that ``it shall be possible for special programs to trigger their own data processing recipes" (DMS-REQ-0320). However, there is currently no requirement for the scheduler to initiate non-standard processing with a certain pipeline if the data was executed under a given program. Since the data processing is scheduler-driven, this would have to be an OSS requirement.
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-2} The purpose of this ticket is to figure out whether and how auto-triggering of alternative data processing recipes for Special Programs can be supported by the DM system. If it can't be, a change request on DMS-REQ-0320 might be needed to down-scope. \end{enumerate}

% % % % % % % % % % % % % % % % % % 
\subsection{Support for User-Driven Processing of Special Programs Data}\label{ssec:dmplans_L3}

As explained above, LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs whenever possible. In Section \ref{ssec:docrev_sizing} we describe how LSST has allocated 10\% of its computational and storage capacity for this processing because 10\% of the images will be generated by Special Programs.

However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3). LSST has furthermore allocated an {\it additional} 10\% of its computing resources for user-driven processing and analysis. This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).

In the case where a Special Program requires an intensive amount of additional computational resources (i.e., too large a fraction of the 10\% allocated), the processing will have to be done on a separate, external computational resource. To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.

In the case where user-driven processing pipelines consume a small amount of that 10\%, data export and processing on external computation resources will not be necessary. Users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their pipelines on LSST servers, monitor their progress, and analyze their output. DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.

However, in the case where user-driven processing pipelines consume a {\bf large} amount of that 10\%, data export and processing on external computation resources {\bf will} be necessary. The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.

{\bf Level 3 Alert Stream? --} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package. K.-T. raises the issue that running servers or publishing VOEvents from DACs may be tricky due to network and security considerations, but that otherwise publishing in non-LSST Alert-like packets ought to be fine.

{\bf Lower Latency for Data Access? --} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner. Furthermore, download limits might be imposed and it isn't clear whether it will be possible for a user to pull over the full nightly amount of data if, e.g., they can fund their own server system to process it.

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-1} The purpose of this ticket is to ask Gregory about whether there are aspects to the Science Platform that need to be developed to better serve the needs of the community doing Special programs processing. Examples include support for Level 3 alert streams and a lower latency for access to processed visit images.  \end{enumerate}

It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team. For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2. An alternative option to ``adopted" code is ``adopted" data products: situations in which Level 3 code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-2} The purpose of this ticket is to clarify whether there is a process in place, and if staffing needs have been defined, for adopting Level 3 code into DM's pipelines, and/or for adopting data products, verifying them, and hosting their public release. This issue is not limited to Special Programs, but is likely to be encountered in that context. \end{enumerate}

% % % % % % % % % % % % % % % % % %
\clearpage
\section{DM Documentation Review}\label{sec:docrev}

Here we gather every requirement or specification relevant to Special Programs in the existing DM documentation, identify areas where the intent of LSST DM is not adequately represented, and suggest clarifications, changes, or additions to these documents.

% % % % % % %
\subsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{ssec:docrev_srd}

% % % % % % %
\subsection{LSST System Requirements, LSR, LSE-29, \cite{LSE-29}}\label{ssec:docrev_lsr}

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-0} RFC (LSE-29): We propose to add a high-level requirement that DM reconfigure its pipelines and produce unique, separate data products for the accepted Special Programs, wherever possible, to better reflect its intent. This would flow-down to the DMSR. \end{enumerate}

$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-1} RFC (LSE-29): We propose to add a requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102). E.g., that data products from reconfigured pipelines will be made available on a intermediate time scale where the science goals require it. \end{enumerate}
